---
title: "A machine learning approach to evaluate exercise performance"
author: "JMW0103"
date: "Thursday, September 17, 2015"
output: html_document
---
Load the required packages for this analysis
```{r warning=FALSE, message=FALSE }
library(digest)
library(caret)
library(randomForest)
set.seed(91295)
```


The correct data sets are required for this analysis:  
1) a set of data >19000 instances from  **https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv**   
2) a set of test instances for _automatic_  __evaluation of the model__  **https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv**

```{r message=FALSE,results='hide'}
library(digest)
if(file.exists("pml-training.csv")&&(identical(digest("pml-training.csv",file=TRUE),"56926c78af383dcdc2060407942e52e9"))){   
} else{  download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv","pml-training.csv")}

if(file.exists("pml-testing.csv")&&(identical(digest("pml-testing.csv",file=TRUE),"bc4174f3ec5dfcc5c570a1d2709272d9"))){   
} else{  download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv","pml-testing.csv") }
```
Manual inspection of the training data set indicates 3 distinct labels for data not available : **"NA", ""** and **"#DIV/0!"**
Read the data sets into data frames, using the above strings to indicate NA in the dataframe.  
```{r}
baseData<-read.csv("pml-training.csv",na.strings=c("NA","#DIV/0!",""))
testData<-read.csv("pml-testing.csv",na.strings=c("NA","#DIV/0!",""))
```

Determine the pre-processing and cleaning required of the data sets:


* Fields 1-7  are irrelevant and can produce spurious overfitting  
        + remove from in the training data

* determine all fields where >5% of entries are NA
* remove these columns from the test and training

```{r}
numInst<-length(baseData[,1])
baseStrip_1<-baseData[,-(1:7)] ; testStrip_1<-testData[,-(1:7)]
na_level<-colSums((is.na(baseStrip_1))/numInst)<0.05
baseStrip_2<-baseStrip_1[,na_level] ; testStrip_2<-testStrip_1[,na_level]
```

There are `r length(baseStrip_2[,1])` instances in the training data   
